---
title: "Getting Started with ggplot2"
author: "Jennifer Favaloro"
date: "May 19, 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      fig.width = 10, fig.height = 6, dpi = 100)
```

# Getting Started with ggplot2

This notebook covers the fundamentals of ggplot2 and demonstrates how to create basic plots with customizations. We'll explore the underlying principles and syntax that make ggplot2 a powerful tool for data visualization.

```{r load-packages}
# Load required libraries
library(tidyverse)
library(ggplot2)
```

## The Grammar of Graphics Concept

ggplot2 is based on the Grammar of Graphics, which breaks down visualizations into semantic components:

1. **Data**: The dataset being plotted
2. **Aesthetics**: Mapping of variables to visual properties (x, y, color, size, etc.)
3. **Geometries**: The actual shapes representing data (points, lines, bars, etc.)
4. **Facets**: For creating small multiples (subplots)
5. **Statistics**: Statistical transformations (counts, means, etc.)
6. **Coordinates**: The coordinate system (Cartesian, polar, etc.)
7. **Themes**: Controlling the display (fonts, colors, backgrounds, etc.)

This layered approach allows for creating complex visualizations by combining simple components.

## Basic ggplot2 Syntax and Structure

Let's first load our dataset and explore the basic syntax of ggplot2.

```{r load-data}
# Load the built-in mtcars dataset
data(mtcars)
head(mtcars)

# Convert cylinders to factor for better visualization
mtcars$cyl <- as.factor(mtcars$cyl)  # Number of cylinders
mtcars$am <- as.factor(mtcars$am)    # Transmission (0 = automatic, 1 = manual)
```

The basic structure of a ggplot2 plot follows this pattern:

```
ggplot(data, aes(x, y)) + geom_function()
```

Let's see what happens when we create an empty plot with just the axes:

```{r empty-plot}
# Empty plot with just the axes
ggplot(mtcars, aes(x = wt, y = mpg))
```

Notice that ggplot2 creates the coordinate system, but there's nothing plotted yet because we haven't specified a geometry. Let's add points to create a scatter plot:

```{r basic-scatter}
# Adding points to create a scatter plot
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```

Now we can see the relationship between car weight and fuel efficiency.

## Creating Basic Plot Types

Let's explore how to create different types of plots using ggplot2.

### Scatter Plot

```{r scatter-plot}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(title = "Car Weight vs. Fuel Efficiency",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon")
```

### Bar Chart

Let's use the built-in diamonds dataset to create a bar chart showing the count of diamonds by cut quality:

```{r bar-chart}
# Using the built-in diamonds dataset
ggplot(diamonds, aes(x = cut)) +
  geom_bar() +
  labs(title = "Count of Diamonds by Cut Quality",
       x = "Cut Quality",
       y = "Count")
```

### Bar Chart with Calculated Values

Sometimes we want to plot pre-calculated values rather than counts. Let's calculate the average MPG by cylinder group and plot it:

```{r bar-chart-calculated}
mtcars_summary <- mtcars %>%
  group_by(cyl) %>%
  summarize(avg_mpg = mean(mpg))

ggplot(mtcars_summary, aes(x = cyl, y = avg_mpg)) +
  geom_col() +
  labs(title = "Average MPG by Number of Cylinders",
       x = "Number of Cylinders",
       y = "Average MPG")
```

### Line Graph

Line graphs are great for showing trends over time. Let's use the economics dataset to plot unemployment over time:

```{r line-graph}
# Using the economics dataset
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line() +
  labs(title = "US Unemployment Over Time",
       x = "Date",
       y = "Unemployment (thousands)")
```

### Histogram

Histograms help visualize the distribution of a continuous variable:

```{r histogram}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 10) +
  labs(title = "Distribution of Fuel Efficiency",
       x = "Miles Per Gallon",
       y = "Count")
```

### Density Plot

Density plots are smoothed versions of histograms, useful for showing the distribution shape:

```{r density-plot}
ggplot(mtcars, aes(x = mpg)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(title = "Density of Fuel Efficiency",
       x = "Miles Per Gallon",
       y = "Density")
```

## Customizing Plot Aesthetics

One of the strengths of ggplot2 is how easily we can map variables to visual properties (aesthetics) like color, size, and shape.

### Adding Color

```{r color-aesthetic}
ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +
  geom_point(size = 3) +
  labs(title = "Car Weight vs. Fuel Efficiency by Cylinders",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon",
       color = "Cylinders")
```

### Adding Shape

```{r shape-aesthetic}
ggplot(mtcars, aes(x = wt, y = mpg, color = cyl, shape = am)) +
  geom_point(size = 3) +
  labs(title = "Car Weight vs. Fuel Efficiency",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon",
       color = "Cylinders",
       shape = "Transmission (0 = auto, 1 = manual)")
```

### Customizing Colors

```{r custom-colors}
ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("red", "green", "blue")) +
  labs(title = "Car Weight vs. Fuel Efficiency by Cylinders",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon",
       color = "Cylinders")
```

### Adding a Smoothing Line

```{r smoothing-line}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Car Weight vs. Fuel Efficiency with Trend Line",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon")
```

### Customizing the Theme

```{r custom-theme}
ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +
  geom_point(size = 3) +
  labs(title = "Car Weight vs. Fuel Efficiency by Cylinders",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon",
       color = "Cylinders") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    legend.position = "bottom"
  )
```

## Saving and Exporting Visualizations

It's often useful to save your visualizations to share them or include them in reports. ggplot2 makes this easy with the `ggsave()` function.

```{r save-plot, eval=FALSE}
# Create a plot to save
p <- ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +
  geom_point(size = 3) +
  labs(title = "Car Weight vs. Fuel Efficiency by Cylinders",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon",
       color = "Cylinders") +
  theme_minimal()

# Save as PNG
ggsave("my_plot.png", p, width = 8, height = 6, dpi = 300)

# Save as PDF
ggsave("my_plot.pdf", p, width = 8, height = 6)

# Save as SVG
ggsave("my_plot.svg", p, width = 8, height = 6)
```

## Summary

In this notebook, we've learned the basics of ggplot2, including:

1. The Grammar of Graphics concept and how it applies to ggplot2
2. Basic ggplot2 syntax and structure
3. Creating different types of plots (scatter plots, bar charts, line graphs, histograms, and density plots)
4. Customizing plot aesthetics (color, shape, size)
5. Adding statistical transformations (smoothing lines)
6. Customizing themes
7. Saving and exporting visualizations

In the next section, we'll explore more advanced ggplot2 techniques like faceting, combining multiple geoms, and using color effectively.
