---
title: "Specialized Visualizations"
author: "Jennifer Favaloro"
date: "May 19, 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      fig.width = 10, fig.height = 6, dpi = 100)
```

# Specialized Visualizations

This notebook demonstrates specialized visualization techniques in R, including statistical visualizations, correlation plots, and heatmaps. These visualizations are particularly useful for exploratory data analysis and statistical inference.

```{r load-packages}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(reshape2)
library(corrplot)
library(patchwork)
```

## Statistical Visualizations

Statistical visualizations help us understand the distribution and summary statistics of our data. Let's explore various statistical visualization techniques.

### Box Plots

Box plots (also called box-and-whisker plots) show the distribution of a continuous variable through quartiles, with outliers plotted as individual points.

```{r boxplots}
# First, let's prepare the mtcars data
data(mtcars)
mtcars$cyl <- as.factor(mtcars$cyl)

# Box plots
ggplot(mtcars, aes(x = cyl, y = mpg, fill = cyl, group = cyl)) +
  geom_boxplot() +
  labs(title = "Fuel Efficiency by Number of Cylinders",
       x = "Number of Cylinders",
       y = "Miles Per Gallon",
       fill = "Cylinders") +
  theme_minimal()
```

In the box plot:
- The box shows the interquartile range (IQR), from the 25th to 75th percentile
- The line inside the box is the median
- The whiskers extend to the smallest and largest values within 1.5 * IQR
- Points outside the whiskers are outliers

### Violin Plots

Violin plots combine aspects of box plots and density plots, showing both the distribution and key statistics.

```{r violin-plots}
# Violin plots
ggplot(mtcars, aes(x = cyl, y = mpg, fill = cyl, group = cyl)) +
  geom_violin(trim = FALSE) +
  labs(title = "Fuel Efficiency Distribution by Number of Cylinders",
       x = "Number of Cylinders",
       y = "Miles Per Gallon",
       fill = "Cylinders") +
  theme_minimal()
```

The width of the violin at each point represents the density of data at that value. This helps visualize the distribution shape.

### Box Plot + Violin Plot Comparison

Let's compare box plots and violin plots side by side:

```{r box-violin-comparison}
# Box plot
p1 <- ggplot(mtcars, aes(x = cyl, y = mpg, fill = cyl, group = cyl)) +
  geom_boxplot() +
  labs(title = "Box Plot",
       x = "Cylinders",
       y = "MPG") +
  theme_minimal() +
  theme(legend.position = "none")

# Violin plot
p2 <- ggplot(mtcars, aes(x = cyl, y = mpg, fill = cyl, group = cyl)) +
  geom_violin(trim = FALSE) +
  labs(title = "Violin Plot",
       x = "Cylinders",
       y = "MPG") +
  theme_minimal() +
  theme(legend.position = "none")

# Using patchwork to combine plots
p1 + p2 + plot_annotation(
  title = "Comparison of Box Plot and Violin Plot",
  subtitle = "Visualizing the same data with different techniques"
)
```

Each type of plot has its strengths:
- Box plots focus on the quartiles and outliers
- Violin plots show the full distribution shape

### Density Plots

Density plots show the distribution of a continuous variable, similar to a smoothed histogram:

```{r density-plots}
# Density plots
ggplot(mtcars, aes(x = mpg, fill = cyl, group = cyl)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density of Fuel Efficiency by Number of Cylinders",
       x = "Miles Per Gallon",
       y = "Density",
       fill = "Cylinders") +
  theme_minimal()
```

### QQ Plots

QQ plots (quantile-quantile plots) help assess if a dataset follows a particular theoretical distribution, typically the normal distribution:

```{r qq-plots}
# QQ plots
ggplot(mtcars, aes(sample = mpg)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "QQ Plot of Fuel Efficiency",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

If the points roughly follow the line, it suggests the data is normally distributed. Deviations from the line indicate departures from normality.

We can also create QQ plots by group:

```{r qq-plots-by-group}
# QQ plots by group
ggplot(mtcars, aes(sample = mpg, color = cyl, group = cyl)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "QQ Plot of Fuel Efficiency by Number of Cylinders",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles",
       color = "Cylinders") +
  theme_minimal()
```

## Visualizing Relationships

Understanding relationships between variables is a key part of data analysis. Let's explore some techniques for visualizing relationships.

### Correlation Matrix

First, let's compute the correlation matrix for numeric variables in the mtcars dataset:

```{r correlation-matrix}
# Select numeric variables for correlation
mtcars_numeric <- mtcars[, c("mpg", "disp", "hp", "drat", "wt", "qsec")]
cor_matrix <- cor(mtcars_numeric)
print(cor_matrix)
```

### Correlation Plot with corrplot

The corrplot package provides a nice visualization of correlation matrices:

```{r corrplot}
# Correlation plot with corrplot
corrplot(cor_matrix, 
         method = "circle", 
         type = "upper", 
         order = "hclust",
         tl.col = "black",
         tl.srt = 45,
         addCoef.col = "black",
         number.cex = 0.7,
         title = "Correlation Matrix of Car Features",
         mar = c(0, 0, 2, 0))
```

In this plot:
- Blue circles indicate positive correlations
- Red circles indicate negative correlations
- The size of the circle represents the strength of the correlation
- The numerical values show the exact correlation coefficients

### Heatmap of Correlation Matrix

We can also visualize the correlation matrix as a heatmap using ggplot2:

```{r correlation-heatmap}
# Melt the correlation matrix for ggplot2
melted_cor <- melt(cor_matrix)

# Create a heatmap
ggplot(melted_cor, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  coord_fixed() +
  labs(title = "Correlation Heatmap of Car Features",
       x = "",
       y = "")
```

### Scatter Plot Matrix

A scatter plot matrix shows pairwise relationships between multiple variables:

```{r scatter-plot-matrix}
# Scatter plot matrix
pairs(mtcars_numeric, 
      main = "Scatter Plot Matrix of Car Features",
      pch = 21,
      bg = rainbow(nrow(mtcars_numeric))[rank(mtcars$cyl)])
```

Each cell in the matrix shows the relationship between the variable on the x-axis and the variable on the y-axis. This is a quick way to explore relationships between multiple variables simultaneously.

## Time Series Visualizations

Time series data is common in many fields. Let's explore techniques for visualizing time series data using the economics dataset from ggplot2.

### Basic Time Series Plot

```{r basic-time-series}
# Load economics dataset
data(economics, package = "ggplot2")

# Basic time series plot
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line(color = "blue") +
  labs(title = "US Unemployment Over Time",
       x = "Date",
       y = "Unemployment (thousands)") +
  theme_minimal()
```

### Multiple Time Series

We often want to compare multiple time series. Let's reshape the data to plot multiple series:

```{r multiple-time-series}
# Prepare data for multiple time series
economics_long <- economics %>%
  select(date, unemploy, psavert) %>%
  rename(Unemployment = unemploy, 
         "Personal Savings Rate" = psavert) %>%
  melt(id.vars = "date", variable.name = "series")

# Plot multiple time series
ggplot(economics_long, aes(x = date, y = value, color = series)) +
  geom_line() +
  labs(title = "US Economic Indicators Over Time",
       x = "Date",
       y = "Value",
       color = "Indicator") +
  theme_minimal() +
  facet_wrap(~ series, scales = "free_y", ncol = 1)
```

Using `facet_wrap` with `scales = "free_y"` allows each series to have its own y-axis scale, which is useful when the series have different units or magnitudes.

### Seasonal Decomposition Visualization

Time series often contain multiple components: trend, seasonality, and random variations. We can visualize these components separately.

For demonstration, let's create a simulated seasonal time series:

```{r simulated-time-series}
# Create a simulated seasonal time series
set.seed(123)
time <- seq(as.Date("2010-01-01"), as.Date("2019-12-31"), by = "month")
trend <- seq(1, length(time)) * 0.1
seasonal <- sin(2 * pi * seq(1, length(time)) / 12) * 5
random <- rnorm(length(time), 0, 1)
value <- trend + seasonal + random

ts_data <- data.frame(
  date = time,
  value = value,
  trend = trend,
  seasonal = seasonal,
  random = random
)

# Plot the components
p1 <- ggplot(ts_data, aes(x = date, y = value)) +
  geom_line() +
  labs(title = "Original Time Series",
       x = "",
       y = "Value") +
  theme_minimal()

p2 <- ggplot(ts_data, aes(x = date, y = trend)) +
  geom_line(color = "blue") +
  labs(title = "Trend Component",
       x = "",
       y = "Trend") +
  theme_minimal()

p3 <- ggplot(ts_data, aes(x = date, y = seasonal)) +
  geom_line(color = "green") +
  labs(title = "Seasonal Component",
       x = "",
       y = "Seasonal") +
  theme_minimal()

p4 <- ggplot(ts_data, aes(x = date, y = random)) +
  geom_line(color = "red") +
  labs(title = "Random Component",
       x = "Date",
       y = "Random") +
  theme_minimal()

# Combine the plots using patchwork
(p1 / p2 / p3 / p4) + 
  plot_annotation(
    title = "Time Series Decomposition",
    subtitle = "Breaking down a time series into its components"
  )
```

This visualization makes it easy to see:
- The original time series with its overall pattern
- The long-term trend component (steadily increasing)
- The seasonal component (repeating pattern)
- The random noise component

## Combining Plots with patchwork

The patchwork package makes it easy to combine multiple plots into a single figure. We've already used it in several examples, but let's explore it more explicitly.

First, let's create some basic plots:

```{r basic-plots-for-patchwork}
# Create some basic plots
p1 <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Weight vs. MPG",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon") +
  theme_minimal()

p2 <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Horsepower vs. MPG",
       x = "Horsepower",
       y = "Miles Per Gallon") +
  theme_minimal()

p3 <- ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 10, fill = "skyblue", color = "black") +
  labs(title = "MPG Distribution",
       x = "Miles Per Gallon",
       y = "Count") +
  theme_minimal()

p4 <- ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl), group = cyl)) +
  geom_boxplot() +
  labs(title = "MPG by Cylinders",
       x = "Cylinders",
       y = "Miles Per Gallon") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Side by Side Arrangement

```{r patchwork-side-by-side}
# Side by side arrangement
p1 + p2
```

### Stacked Arrangement

```{r patchwork-stacked}
# Stacked arrangement
p1 / p2
```

### Grid Arrangement

```{r patchwork-grid}
# Grid arrangement
(p1 + p2) / (p3 + p4)
```

### Complex Layout

```{r patchwork-complex}
# Complex layout
layout <- "
AABB
AACC
"
p1 + p2 + p3 + plot_layout(design = layout)
```

### With Annotations

```{r patchwork-with-annotations}
# With annotations
(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Car Performance Analysis Dashboard",
    subtitle = "Multiple visualizations of the mtcars dataset",
    caption = "Source: 1974 Motor Trend US magazine",
    tag_levels = "A"
  )
```

The patchwork package provides a flexible and intuitive way to create complex layouts of multiple plots. This is especially useful for creating dashboards or combining related visualizations.

## Summary

In this notebook, we've explored specialized visualization techniques including:

1. Statistical visualizations (box plots, violin plots, density plots, QQ plots)
2. Relationship visualizations (correlation matrices, heatmaps, scatter plot matrices)
3. Time series visualizations (basic time series, multiple time series, seasonal decomposition)
4. Combining plots with patchwork

These techniques provide powerful tools for exploring and communicating patterns in your data. In the next section, we'll explore interactive visualizations using plotly.
