---
title: "Interactive Visualizations with Plotly"
author: "Jennifer Favaloro"
date: "May 19, 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      fig.width = 10, fig.height = 6, dpi = 100)
```

# Interactive Visualizations with Plotly

This notebook introduces interactive visualizations using the plotly package for R. Plotly creates interactive web-based visualizations that allow users to hover, zoom, pan, and select data points. These interactive features can greatly enhance data exploration and presentation.

```{r load-packages}
# Load required libraries
library(plotly)
library(dplyr)
library(reshape2)
```

## Example 1: Interactive Scatter Plot

Let's start with a basic interactive scatter plot using the built-in mtcars dataset:

```{r scatter-plot}
# Using the built-in mtcars dataset
scatter_plot <- plot_ly(data = mtcars, 
                        x = ~wt, 
                        y = ~mpg, 
                        color = ~factor(cyl),
                        size = ~hp,
                        text = ~paste("Car: ", rownames(mtcars), 
                                     "<br>Weight: ", wt, 
                                     "<br>MPG: ", mpg,
                                     "<br>Cylinders: ", cyl,
                                     "<br>Horsepower: ", hp),
                        hoverinfo = "text",
                        type = "scatter",
                        mode = "markers") %>%
  layout(title = "Car Weight vs. Fuel Efficiency",
         xaxis = list(title = "Weight (1000 lbs)"),
         yaxis = list(title = "Miles Per Gallon"),
         legend = list(title = list(text = "Cylinders")))

# Display the plot
scatter_plot
```

In this interactive plot:
- Hover over points to see detailed information
- Click and drag to zoom into a specific region
- Double-click to reset the zoom
- Click on legend items to hide/show categories
- Use the mode bar (top-right) for additional tools

Key features of this plotly scatter plot:
- Points are colored by the number of cylinders
- Point size represents horsepower
- Custom hover text shows multiple variables
- The layout is customized with titles and legend settings

## Example 2: Bar Chart

Next, let's create an interactive bar chart with custom colors:

```{r bar-chart}
# Summarize mtcars data by cylinder
cyl_summary <- mtcars %>%
  group_by(cyl) %>%
  summarize(avg_mpg = mean(mpg),
            count = n())

# Create a bar chart with custom colors
bar_plot <- plot_ly(data = cyl_summary,
                    x = ~factor(cyl),
                    y = ~avg_mpg,
                    type = "bar",
                    marker = list(color = c("#1f77b4", "#ff7f0e", "#2ca02c")),
                    text = ~paste("Cylinders: ", cyl,
                                 "<br>Average MPG: ", round(avg_mpg, 1),
                                 "<br>Count: ", count),
                    hoverinfo = "text") %>%
  layout(title = "Average Fuel Efficiency by Number of Cylinders",
         xaxis = list(title = "Number of Cylinders"),
         yaxis = list(title = "Average MPG"))

# Display the plot
bar_plot
```

This bar chart shows the average MPG for each cylinder category, with hover text that includes the average MPG and the number of cars in each category.

## Example 3: Line Chart with Multiple Series

Line charts are particularly useful for time series data. Let's create a line chart with multiple series using the economics data:

```{r line-chart}
# Load the economics dataset from ggplot2
data(economics, package = "ggplot2")

# Create a subset of the data
econ_subset <- tail(economics, 120)  # Last 10 years of data

# Create a line chart with multiple series
line_plot <- plot_ly(data = econ_subset, x = ~date) %>%
  add_trace(y = ~unemploy, name = "Unemployment",
            type = "scatter", mode = "lines",
            line = list(color = "#1f77b4", width = 2)) %>%
  add_trace(y = ~uempmed, name = "Median Unemployment Duration",
            type = "scatter", mode = "lines",
            line = list(color = "#ff7f0e", width = 2)) %>%
  layout(title = "US Unemployment Trends",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Value"),
         legend = list(x = 0.1, y = 0.9))

# Display the plot
line_plot
```

This line chart shows two unemployment metrics over time:
- Total unemployment (in thousands of people)
- Median duration of unemployment (in weeks)

The interactive nature allows users to zoom into specific time periods and compare the two metrics.

## Example 4: Heatmap

Heatmaps are useful for visualizing matrices of data, such as correlation matrices:

```{r heatmap}
# Calculate correlation matrix
cor_matrix <- cor(mtcars[, c("mpg", "disp", "hp", "drat", "wt", "qsec")])

# Convert to a format suitable for plotly
cor_data <- as.data.frame(cor_matrix)
cor_data$variable1 <- rownames(cor_data)
cor_long <- tidyr::pivot_longer(cor_data, 
                               cols = c("mpg", "disp", "hp", "drat", "wt", "qsec"),
                               names_to = "variable2", 
                               values_to = "correlation")

# Create a heatmap
heatmap_plot <- plot_ly(data = cor_long,
                        x = ~variable1,
                        y = ~variable2,
                        z = ~correlation,
                        type = "heatmap",
                        colorscale = "RdBu",
                        zmin = -1, zmax = 1,
                        text = ~paste("Variables: ", variable1, " & ", variable2,
                                     "<br>Correlation: ", round(correlation, 2)),
                        hoverinfo = "text") %>%
  layout(title = "Correlation Matrix of Car Features",
         xaxis = list(title = ""),
         yaxis = list(title = ""))

# Display the plot
heatmap_plot
```

This heatmap shows the correlations between different car features, with blue representing positive correlations and red representing negative correlations. The interactive hover text shows the exact correlation value.

## Example 5: 3D Scatter Plot

Plotly makes it easy to create interactive 3D visualizations:

```{r scatter-3d}
# Create a 3D scatter plot
scatter_3d <- plot_ly(data = mtcars,
                      x = ~wt,
                      y = ~hp,
                      z = ~mpg,
                      color = ~factor(cyl),
                      size = ~qsec,
                      text = ~paste("Car: ", rownames(mtcars),
                                   "<br>Weight: ", wt,
                                   "<br>Horsepower: ", hp,
                                   "<br>MPG: ", mpg,
                                   "<br>Cylinders: ", cyl),
                      hoverinfo = "text",
                      type = "scatter3d",
                      mode = "markers") %>%
  layout(title = "3D Visualization of Car Features",
         scene = list(
           xaxis = list(title = "Weight (1000 lbs)"),
           yaxis = list(title = "Horsepower"),
           zaxis = list(title = "Miles Per Gallon")
         ))

# Display the plot
scatter_3d
```

This 3D scatter plot visualizes the relationship between weight, horsepower, and MPG, with points colored by the number of cylinders and sized by quarter-mile time. 

The interactive features of 3D plotly plots are particularly powerful:
- Click and drag to rotate the plot
- Scroll to zoom in and out
- Double-click to reset the view
- Hover for details about each point

## Example 6: Box Plot

Box plots are useful for comparing distributions across categories:

```{r box-plot}
# Create a box plot
box_plot <- plot_ly(data = mtcars,
                    x = ~factor(cyl),
                    y = ~mpg,
                    color = ~factor(cyl),
                    type = "box",
                    boxpoints = "all",
                    jitter = 0.3,
                    pointpos = 0) %>%
  layout(title = "Distribution of Fuel Efficiency by Cylinder Count",
         xaxis = list(title = "Number of Cylinders"),
         yaxis = list(title = "Miles Per Gallon"),
         showlegend = FALSE)

# Display the plot
box_plot
```

This interactive box plot shows the distribution of MPG for each cylinder category, including all individual data points. The `boxpoints = "all"` setting ensures that all individual points are shown, and `jitter = 0.3` adds some horizontal random noise to prevent overlapping.

## Example 7: Histogram with Multiple Traces

We can create overlapping histograms to compare distributions:

```{r histogram}
# Create overlapping histograms
hist_plot <- plot_ly(alpha = 0.6) %>%
  add_histogram(x = mtcars$mpg[mtcars$cyl == 4],
                name = "4 Cylinders",
                marker = list(color = "#1f77b4")) %>%
  add_histogram(x = mtcars$mpg[mtcars$cyl == 6],
                name = "6 Cylinders",
                marker = list(color = "#ff7f0e")) %>%
  add_histogram(x = mtcars$mpg[mtcars$cyl == 8],
                name = "8 Cylinders",
                marker = list(color = "#2ca02c")) %>%
  layout(title = "Distribution of MPG by Cylinder Count",
         xaxis = list(title = "Miles Per Gallon"),
         yaxis = list(title = "Count"),
         barmode = "overlay")

# Display the plot
hist_plot
```

This plot shows the distribution of MPG for each cylinder category as overlapping histograms. The `barmode = "overlay"` setting ensures that the histograms are overlaid rather than stacked.

## Example 8: Bubble Chart

A bubble chart is a scatter plot where a third variable is represented by the size of the bubbles:

```{r bubble-chart}
# Create a bubble chart
bubble_plot <- plot_ly(data = mtcars,
                       x = ~wt,
                       y = ~mpg,
                       size = ~disp,
                       color = ~factor(am),
                       colors = c("#1f77b4", "#ff7f0e"),
                       text = ~paste("Car: ", rownames(mtcars),
                                    "<br>Weight: ", wt,
                                    "<br>MPG: ", mpg,
                                    "<br>Displacement: ", disp,
                                    "<br>Transmission: ", ifelse(am == 0, "Automatic", "Manual")),
                       hoverinfo = "text",
                       type = "scatter",
                       mode = "markers",
                       marker = list(opacity = 0.7, sizemode = "diameter")) %>%
  layout(title = "Car Weight vs. Fuel Efficiency (Bubble Size = Displacement)",
         xaxis = list(title = "Weight (1000 lbs)"),
         yaxis = list(title = "Miles Per Gallon"),
         legend = list(title = list(text = "Transmission")))

# Display the plot
bubble_plot
```

In this bubble chart:
- X-axis represents car weight
- Y-axis represents fuel efficiency (MPG)
- Bubble size represents engine displacement
- Color represents transmission type (automatic or manual)

## Saving Plotly Visualizations

Plotly visualizations can be saved as interactive HTML files for sharing or embedding in web pages:

```{r save-plotly, eval=FALSE}
# Example of saving a plot to an HTML file
htmlwidgets::saveWidget(as_widget(scatter_plot), "scatter_plot.html")

# You can also export to static image formats
# orca(scatter_plot, "scatter_plot.png")
# Note: This requires the orca package to be installed
# install.packages("processx")
# plotly::orca_install()
```

## Summary

In this notebook, we've explored interactive data visualization with plotly, including:

1. Interactive scatter plots with hover information
2. Bar charts with custom colors and hover text
3. Line charts with multiple series
4. Interactive heatmaps for correlation matrices
5. 3D scatter plots for exploring multidimensional relationships
6. Box plots with jittered points
7. Overlapping histograms for comparing distributions
8. Bubble charts for visualizing multiple variables

The interactive nature of plotly visualizations makes them particularly useful for:
- Exploratory data analysis
- Interactive dashboards
- Web-based reports
- Presentations where you want to show both the big picture and the details

In the next section, we'll explore mosaic plots for visualizing categorical data relationships.
